\begin{acknowledgements}
  I am indebted to a great many people for helping me through these five
  years.

  First, I want to thank Ranjit Jhala for being such a great advisor.
  He has always been supremely supportive and is brimming with ideas for
  interesting projects, even when you feel like you have none of your
  own. He also has a brilliant ability to force you to keep trying
  (usually failing) to explain your ideas, until finally he decides to
  step in and summarize what you've been struggling to say with the
  utmost clarity and precision. Finally, I want to thank Ranjit for
  pushing me to investigate using machine learning to predict the source
  of type errors; I was quite reluctant at first, but it turned out to
  be a very enlightening and productive direction!

  I also thank Bill Griswold, Philip Guo, Jim Hollan, and Sorin Lerner
  for serving on my committee. In particular, I want to thank Jim for
  inviting me to his research group meetings and Bill for reading so
  many papers with me in my early years here. Thanks also to Ingolf
  Krueger, who enabled this great experience by recruiting me to UC
  San Diego in the first place.

  Next, I want to thank all of my collaborators. Wes Weimer seems to
  have an endless supply of insightful questions that helped guide my
  research, and he also provided invaluable assistance in designing and
  running our user studies. Kamalika Chaudhuri and Huma Sibghat helped
  me quickly get up to speed in the foreign field of machine learning,
  and ensured that our models and experiments were on sound footing.

  I want to thank my many labmates in the PL group at UCSD for making my
  time here so enjoyable. In particular, I thank Alexander Bakst,
  Valentin Robert, Zach Tatlock, and Niki Vazou for welcoming me into
  the group and providing a supportive environment.

  I also want to thank Gabrielle Allen and Douglas Troeger for being
  early mentors during my undergraduate career. Gab invited me to work
  with her research group at LSU, which made me realize that I actually
  do enjoy programming when I have a purpose, and lead to three
  publications before I even started grad school. Dr.\ Troeger inspired
  and nurtured my interest in functional programming with his
  Programming Paradigms course at CCNY.\@ I surely wouldn't have even
  applied to grad school without them.

  Last but not least, I want to thank my family for being supportive
  throughout this whole experience. Grad school can be very difficult at
  times, to put it lightly, but my parents, my sister, and my wife Megan
  never wavered in their support and confidence that I could persevere.

  \paragraph{Published Materials Adapted for this Dissertation}
  The dissertation author was the primary investigator on several papers
  that have either been published or are in submission, which have been
  adapted for this dissertation.
  %
  The following publications include material that has been incorporated
  into \autoref{chp:intro} and \autoref{chp:nate}:
  %
  \begin{itemize}
  \item \fullcite{Seidel2017-pj}
  \end{itemize}
  %
  The following publications include material that has been incorporated
  into \autoref{chp:intro} and \autoref{chp:nanomaly}:
  %
  \begin{itemize}
  \item \fullcite{Seidel2016-ul}
  \item \fullcite{Seidel2017-gw}
  \end{itemize}
\end{acknowledgements}


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
